---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Loading Data from zip file.

```{r, echo=TRUE}
activity_data <- read.table(unz("./activity.zip", "activity.csv"), header=T, sep=",", colClasses = c("integer","Date", "integer"))
head(activity_data)
```


## What is mean total number of steps taken per day?

Caclculating Total number of steps taken per day.

```{r, echo=TRUE}
steps_per_day <- tapply(activity_data$steps, activity_data$date, sum)
steps_per_day
```

Removing Days having steps 'Not Available'.

```{r, echo=TRUE}
steps_per_day <- steps_per_day[!is.na(steps_per_day)]
steps_per_day
```

Converting to DataFrame.

```{r, echo=TRUE}
steps_per_day_df <- data.frame(Day=names(steps_per_day), Steps=steps_per_day, row.names=NULL)
head(steps_per_day_df)
```

Plotting Histogram.

```{r, echo=TRUE}
library(ggplot2)
ggplot(steps_per_day_df, aes(x = Steps, y = Day)) +
       geom_histogram(stat = 'identity') + coord_flip() + theme(axis.text.x = element_text(angle = 90))
```
Mean of the total nuumber of steps taken per day.

```{r, echo=TRUE}
mean(steps_per_day)
```

Median of the total number of steps taken per day.

```{r, echo=TRUE}
median(steps_per_day)
```

## What is the average daily activity pattern?

Calculating Average across all days for each interval.

```{r, echo=TRUE}
avg_per_interval <- tapply(activity_data$steps, activity_data$interval, mean, na.rm=TRUE)
avg_per_interval
```


Converting to DataFrame.

```{r, echo=TRUE}
avg_per_interval_df <- data.frame(Interval=as.numeric(names(avg_per_interval)), Steps=avg_per_interval, row.names=NULL)
head(avg_per_interval_df)
```

Time Series Plot.

```{r, echo=TRUE}
ggplot(avg_per_interval_df, aes(x=Interval, y=Steps, group=1)) + geom_line() + scale_x_continuous(breaks=seq(-200,2400, 200))
```

Interval With maximum number of steps.

```{r, echo=TRUE}
avg_per_interval_df[which.max(avg_per_interval_df$Steps),]
```


## Imputing missing values

Total number of missing values in the dataset.

```{r, echo=TRUE}
nrow(activity_data[is.na(activity_data$steps), ])
```

Creating new dataset by filling missing values in original dataset.
Missing Values will be filled by mean of steps with respect to interval (calculated above).

```{r, echo=TRUE}

activity_data_filled = activity_data

na_indices <- which(is.na(activity_data$steps))

for (index in na_indices) {
    
    curr_interval <- activity_data[index,]$interval
        
    avg_steps <- avg_per_interval_df[avg_per_interval_df$Interval == curr_interval,]$Steps

    activity_data_filled[index,]$steps <- avg_steps
}

head(activity_data_filled)
```

Caclculating Total number of steps taken per day after filling missing values.

```{r, echo=TRUE}
steps_per_day_filled <- tapply(activity_data_filled$steps, activity_data_filled$date, sum)
steps_per_day_filled
```

Converting to DataFrame.

```{r, echo=TRUE}
steps_per_day_filled_df <- data.frame(Day=names(steps_per_day_filled), Steps=steps_per_day_filled, row.names=NULL)
head(steps_per_day_filled_df)
```

Plotting Histogram.

```{r, echo=TRUE}
library(ggplot2)
ggplot(steps_per_day_filled_df, aes(x = Steps, y = Day)) +
       geom_histogram(stat = 'identity') + coord_flip() + theme(axis.text.x = element_text(angle = 90))
```

Mean of the total nuumber of steps taken per day.

```{r, echo=TRUE}
mean(steps_per_day_filled)
```

Median of the total number of steps taken per day.

```{r, echo=TRUE}
median(steps_per_day_filled)
```

After imputing missing values mean and median values of the Total Steps per day became equal.

## Are there differences in activity patterns between weekdays and weekends?

Creating a new variable to indicate whether given day is weekend or weekday

```{r, echo=TRUE}
library(chron)

daytype <- is.weekend(activity_data_filled$date)
activity_data_filled$daytype <- factor(ifelse(daytype, "weekend", "weekday"))
head(activity_data_filled)

```

Calculating Average across all weekdays and weekends for each interval.

```{r, echo=TRUE}
library(reshape)

avg_per_interval_daytype <- tapply(activity_data_filled$steps, list(activity_data_filled$interval, activity_data_filled$daytype), mean)
avg_per_interval_daytype_df <- as.data.frame(avg_per_interval_daytype)
avg_per_interval_daytype_df$interval <- row.names(avg_per_interval_daytype_df)
avg_per_interval_daytype_df <- melt.data.frame(avg_per_interval_daytype_df)
colnames(avg_per_interval_daytype_df) <- c("interval", "daytype", "steps")
head(avg_per_interval_daytype_df)
```

Plotting Time Series graph with respect to day type.

```{r, echo=TRUE}
ggplot(avg_per_interval_daytype_df, aes(x=as.numeric(interval), y=steps, group=1)) + geom_line() + facet_grid(. ~ daytype, scales="free") + xlab("interval")
```
